---
title: "Preparing covariates for admove"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Preparing covariates for admove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: admove.bib
---

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.show = "hold",
  fig.width = 8,
  fig.height = 7,
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to prepare **gridded covariates** for use with
**admove**. In particular, it demonstrates how common data structures (matrices,
arrays, lists, and raster objects) can be converted into the standard format
expected by `prep_cov()`.

In *admove*, covariates (e.g., habitat suitability, temperature, chlorophyll,
depth) are represented as **fields on a regular spatial grid**, optionally
varying over time. The helper `prep_cov()` standardises different input formats
into a single representation used internally by the package.

`prep_cov()` is designed to accept:

- a numeric **matrix** (one spatial field, single time slice),
- a numeric **array** (multiple time slices),
- a **list** of matrices/arrays (one element per time slice),
- a `RasterLayer` / `RasterBrick` / `RasterStack` (from **raster**, if available),
- a `SpatRaster` (from **terra**, if available).

The output is an object of class `admove_cov` (internally a numeric array) with
dimensions corresponding to space and time.

```{r}
library(admove)
```

# Target data format for all covariates

The standard representation used by *admove* is a 3D array

- `cov[x, y, t]`

where

- the first dimension indexes **x** (east–west),
- the second dimension indexes **y** (north–south),
- the third dimension indexes **time**.

To support plotting and model setup, it is strongly recommended to provide
**dimnames** for the first two dimensions (and ideally also for time):

- `dimnames(cov)[[1]]`: x-coordinates (cell midpoints),
- `dimnames(cov)[[2]]`: y-coordinates (cell midpoints),
- `dimnames(cov)[[3]]`: time labels (preferably numeric times or date strings).

## How time slices are interpreted

The time vector describes **the start of the interval** over which each time
slice should be used and corresponds in *admove* convention to the time since a
origin specified in the time reference information (`admove_tref`). For example,
if the time labels are `0` and `1`, the slices are valid for the first and
second time steps. To which date `0` corresponds to is specified in
`tref()$origin` and the time step between `0` and `1` is specified in
`tref()$units`.

More on that below, for now, we can work with dates, e.g. `2025-01-01` and
`2025-04-01`, or decimal years, e.g. `2025.00` and `2025.25`. In this case,
*admove* uses the first covariate field up to (but not including) `2025.25`, and
the second field from `2025.25` onward. More generally: each slice is assumed
valid from its start time until the start time of the next slice.

# A minimal example in target format

Below we simulate an array of four covariate fields in the target format.

```{r}
## Set seed for reproducibility
set.seed(1)

## Define grid midpoints
x <- seq(-42.5, 42.5, by = 5)
y <- seq(-72.5, 72.5, by = 5)

## Define start times of time intervals
tvals <- c(0, 1, 2, 3)

## Simulate two smooth "basis" fields
mat1 <- outer(x, y, function(xx, yy) {
  exp(-((xx - 10)^2 + (yy + 5)^2) / (2 * 15^2))
}) + matrix(rnorm(length(x) * length(y), 0, 0.02),
            nrow = length(x), ncol = length(y))

mat2 <- outer(x, y, function(xx, yy) {
  exp(-((xx + 20)^2 + (yy - 20)^2) / (2 * 15^2))
}) + matrix(rnorm(length(x) * length(y), 0, 0.02),
            nrow = length(x), ncol = length(y))

## Build a time-varying covariate array as mixtures of mat1/mat2
arr <- array(NA_real_, dim = c(length(x), length(y), length(tvals)))
for (k in seq_along(tvals)) {
  p <- runif(2)
  arr[, , k] <- p[1] * mat1 + p[2] * mat2 + 0.2 * sin(2 * pi * (k - 1) / length(tvals))
}

## Attach dimnames (note: dimnames are character vectors by definition)
dimnames(arr) <- list(
  x = sprintf("%.2f", x),
  y = sprintf("%.2f", y),
  t = sprintf("%.2f", tvals)
)

## Plot two time slices
par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
image(x, y, arr[, , 1], main = paste("t =", dimnames(arr)[[3]][1]), xlab = "x", ylab = "y")
image(x, y, arr[, , 2], main = paste("t =", dimnames(arr)[[3]][2]), xlab = "x", ylab = "y")
par(mfrow = c(1, 1))
```

The array format above is the *target*. In practice, however, covariate data
often arrive in other layouts. The sections below show how to use `prep_cov()`
to standardise common input formats.

# Common format A: a single field as a matrix

Some covariates come as a single field that is effectively constant over time,
such as bathymetry or (static) habitat classifications. We use `mat1` as an
example:

```{r}
image(x, y, mat1, main = "A single matrix (one time slice)", xlab = "x", ylab = "y")
```

You can convert a single matrix using `prep_cov()`:

```{r}
cov1 <- prep_cov(mat1)
```

As the messages indicate, if `mat1` has no row/column names, `prep_cov()` must
fall back to row/column indices, which is rarely what you want. In almost all
real workflows you should use the `x_centers` and `y_centers` arguments of the
function:

```{r}
cov1 <- prep_cov(mat1, x_centers = x, y_centers = y)
```

or attach the grid coordinates via `rownames()` and `colnames()`:

```{r}
rownames(mat1) <- x
colnames(mat1) <- y

cov1 <- prep_cov(mat1)
```

A 2D matrix is automatically promoted to a 3D array with a singleton time
dimension (i.e., `t = 0`):

```{r}
dimnames(cov1)[[3]]
```


# Common format B: multiple time slices as a 3D array

Time-varying covariates (e.g., temperature) are often provided as a 3D array
whose third dimension is time. This already matches the target representation,
so conversion is straightforward:

```{r}
cov2 <- prep_cov(arr)
summary(cov2)
```

The third-dimension labels should describe the start time of each interval
(e.g., decimal years or date strings). See the section *Converting dates to
decimal dates* below for convenient date parsing.

```{r}
dimnames(cov2)[[3]]
```

## Common format C: multiple time slices as a list of matrices

Another common representation is a list where each element corresponds to one
time slice. Here we create such a list from the array above:

```{r}
lst <- lapply(seq_len(dim(arr)[3]), function(k) arr[, , k, drop = FALSE])
cov3 <- prep_cov(lst)
summary(cov3)
```

Internally, the list is converted to a 3D array. If list elements do not have
dimnames, add them before conversion so that coordinates and time metadata are
carried through correctly.

# Common format D: Raster* objects (raster)

Many covariate products are distributed as rasters. If the **raster** package
is available, `prep_cov()` supports `RasterLayer` objects directly
[@hijmans_raster_2025]. This section shows a small example using the simulated
data above.

```{r raster-install, eval=FALSE}
install.packages("raster")
```

```{r raster-create, eval=requireNamespace("raster", quietly=TRUE)}
library(raster)

dx <- diff(x)[1]
dy <- diff(y)[1]

## Define raster extent using cell edges (centers +/- half a cell)
r <- raster::raster(
  xmn = min(x) - dx/2, xmx = max(x) + dx/2,
  ymn = min(y) - dy/2, ymx = max(y) + dy/2,
  res = c(dx, dy),
  crs = NA
)

stopifnot(ncol(r) == length(x), nrow(r) == length(y))
```

## Orientation note (matrix vs raster)

A frequent source of mistakes is that **raster objects and R matrices use
different conventions** for how the y-axis is oriented when values are stored as
a vector. To ensure the raster matches the matrix plot you see with `image()`,
we flip the matrix in the y-direction before assigning values:

```{r raster-fill, eval=requireNamespace("raster", quietly=TRUE)}
mat_topdown <- arr[, , 1][, ncol(arr[, , 1]):1, drop = FALSE]
raster::values(r) <- as.vector(mat_topdown)

par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
image(x, y, arr[, , 1], main = "Standard matrix (image)", xlab = "x", ylab = "y")
plot(r, main = "RasterLayer (raster)")
par(mfrow = c(1, 1))
```

Now the raster can be converted directly:

```{r raster-prep, eval=requireNamespace("raster", quietly=TRUE)}
cov_r <- prep_cov(r)
summary(cov_r)
```

If you prefer not to depend on **raster** inside your workflow, you can always
convert raster-like objects outside *admove* and pass a matrix/array to
`prep_cov()`.

# Common format E: terra data (SpatRaster)

The **terra** package is increasingly common for raster data (GeoTIFF, NetCDF,
etc.) [@hijmans_terra_2026]. If **terra** is available, *admove* can convert
`SpatRaster` objects directly.

```{r terra-install, eval=FALSE}
install.packages("terra")
```

```{r terra-create, eval=requireNamespace("terra", quietly=TRUE)}
library(terra)

dx <- diff(x)[1]
dy <- diff(y)[1]

t <- terra::rast(
  ncols = length(x),
  nrows = length(y),
  xmin  = min(x) - dx/2, xmax = max(x) + dx/2,
  ymin  = min(y) - dy/2, ymax = max(y) + dy/2,
  crs   = ""
)

## Use the same orientation adjustment as in the raster example
mat_topdown <- arr[, , 1][, ncol(arr[, , 1]):1, drop = FALSE]
terra::values(t) <- as.vector(mat_topdown)

plot(t, main = "terra SpatRaster")
```

```{r terra-prep, eval=requireNamespace("terra", quietly=TRUE)}
cov_t <- prep_cov(t)
summary(cov_t)
```

## terra time stacks (multi-layer SpatRaster)

Time-varying fields can be represented as multi-layer `SpatRaster` objects (one
layer per time slice). These can also be converted by `prep_cov()`.

```{r terra-time, eval=requireNamespace("terra", quietly=TRUE)}
t_stack <- c(t, t)

## Layer names should ideally encode time information
names(t_stack) <- c("2025.00", "2025.50")

cov_t_stack <- prep_cov(t_stack, date_decimal = TRUE)
summary(cov_t_stack)
```

# Common format F: NetCDF data (ncdf4)

Many oceanographic and atmospheric products are distributed as NetCDF files.
These often store variables as arrays with dimensions such as `lon × lat × time`
(or `x × y × time`). `prep_cov()` does not read NetCDF files directly, but you
can read a variable with **ncdf4** and then pass a matrix/array to `prep_cov()`
[@pierce_ncdf4_2025].

Below we create a small dummy NetCDF file to demonstrate the workflow.

```{r netcdf-install, eval=FALSE}
install.packages("ncdf4")
```

```{r netcdf-create, eval=requireNamespace("ncdf4", quietly=TRUE)}
library(ncdf4)

tmpfile <- tempfile(fileext = ".nc")

## Define time stamps as Dates (start of each interval)
t_dates <- as.Date(c("2025-01-01", "2025-04-01", "2025-07-01", "2025-10-01"))

## NetCDF dimensions
dim_x <- ncdim_def("x", "km", vals = x)
dim_y <- ncdim_def("y", "km", vals = y)

## NetCDF time is usually numeric with a units string
time_units <- "days since 1970-01-01"
time_vals  <- as.numeric(t_dates - as.Date("1970-01-01"))
dim_t <- ncdim_def("time", time_units, vals = time_vals)

var_cov <- ncvar_def(
  name = "cov",
  units = "1",
  dim = list(dim_x, dim_y, dim_t),
  missval = NA_real_,
  longname = "dummy covariate"
)

nc <- nc_create(tmpfile, vars = list(var_cov))
ncvar_put(nc, var_cov, arr)
nc_close(nc)
```

Now we read the file, extract coordinates and time metadata, and attach dimnames
in the *admove* convention `[x, y, t]`:

```{r netcdf-read, eval=requireNamespace("ncdf4", quietly=TRUE)}
nc <- nc_open(tmpfile)

x_read    <- ncvar_get(nc, "x")
y_read    <- ncvar_get(nc, "y")
time_read <- ncvar_get(nc, "time")  ## numeric days since 1970-01-01
cov_read  <- ncvar_get(nc, "cov")

nc_close(nc)

t_read <- as.Date(time_read, origin = "1970-01-01")

dimnames(cov_read) <- list(
  x = sprintf("%.2f", x_read),
  y = sprintf("%.2f", y_read),
  t = format(t_read)
)

cov_nc <- prep_cov(cov_read, date_format = "%Y-%m-%d")
summary(cov_nc)
```

## Notes on real NetCDF files

Real products often use different dimension names and orders, e.g.
`lon × lat × time`, `time × lat × lon`, or include additional dimensions such as
depth. In those cases you typically need to:

- identify which dimensions correspond to x/lon and y/lat,
- select/aggregate over depth (if present),
- **permute** the array into `[x, y, t]` using `aperm()`.

For example, if the array were `[time, y, x]`, you would do:

```{r, eval=FALSE}
cov_xy_t <- aperm(cov_read, c(3, 2, 1))  ## (time, y, x) -> (x, y, time)
```

and then attach `dimnames(cov_xy_t) <- list(x=..., y=..., t=...)` before calling
`prep_cov()` or use the arguments `x_centers`, `y_centers` and `times`.


# Other functionality of `prep_cov()`

## Converting date strings and decimal years to numeric time

If the third dimension uses date strings, `prep_cov()` can parse them via
`date_format` and `date_origin` and convert them to a numeric time scale (e.g.,
weeks since origin) used internally.

For example, monthly fields labelled as `"YYYY-MM-DD"`:

```{r}
months <- c("2020-01-01", "2020-04-01", "2020-07-01", "2020-10-01")

arr_month <- array(NA_real_, dim = c(length(x), length(y), length(months)))
for (k in seq_along(months)) {
  arr_month[, , k] <- mat1 + 0.1 * k
}

dimnames(arr_month) <- list(
  x = sprintf("%.2f", x),
  y = sprintf("%.2f", y),
  t = months
)

cov_month <- prep_cov(arr_month, date_format = "%Y-%m-%d")
dimnames(cov_month)[[3]]
```

## Specifying time reference information

The time reference information is important as it stores the origin to which
time `0` corresponds to and the units that defines if `1` corresponds to one
day, week, month, year, etc.

For that reason, `prep_cov()` includes the `tref` argument that allows
specifying these important variables. For example, we can prepare the
covariate fields from above with the corresponding `tref` information by

```{r}
cov2 <- prep_cov(arr,
                 tref = list(origin = "2020-01-01",
                            units = "quarter"))
```

Now, the numeric values in `dimnames(cov2)[[3]]` cannot be misinterpreted any
longer as the time reference information is clear:

```{r}
tref(cov2)
```

This information is also provided in the summary:

```{r}
summary(cov2)
```

Of course, the information can also be added after the preparation of the
covariate fields:

```{r}
cov2 <- prep_cov(arr)

cov2 <- add_tref(cov2, list(origin = "2020-01-01",
                              units = "quarter"))

tref(cov2)
```

If dates or decimal years are provided rather than a numeric value, `prep_cov()`
will try to infer the time reference information. For example, for the `cov_nc`
and `cov_t_stack` the information was inferred from the data:

```{r, eval=FALSE}
tref(cov_nc)
tref(cov_t_stack)
```

Nevertheless, we recommend to specify `tref` for covariate fields explicitely.

If the `tref` information was not inferred correctly, the `shift_tref()` and
`scale_tref()` functions allow to correct the information:

```{r}
cov2 <- shift_tref(cov2, origin = "2024-01-01")

cov2 <- scale_tref(cov2, units = "month")
```

Now, the covariate fields corresponds to months starting in January 1st 2024:

```{r}
tref(cov2)
```


## Specifying space reference information

Similar to the time reference information, the spatial information of the
gridded covariate fields also refers to a certain coordinate system (CRS) and
units. Therefore, the covariate fields also contain a `sref` object. However,
in contrast to the time reference information, there is no way to infer the CRS
and unit from the position information alone. Thus, if not specified (default),
the information is just `NA`:

```{r}
sref(cov2)
```

Therefore, it is recommended to specify the spatial reference system by either
using the `sref` argument of `prep_cov()`:

```{r}
cov2 <- prep_cov(arr,
                 tref = list(origin = "2020-01-01",
                            units = "quarter"),
                 sref = list(crs = 4326))

sref(cov2)
```

or by adding the `sref` afterwards:

```{r}
cov2 <- prep_cov(arr,
                 tref = list(origin = "2020-01-01",
                            units = "quarter"))

cov2 <- add_sref(cov2, list(crs = 4326))

sref(cov2)
```



# Sanity checks before modelling

Covariate fields are easy to accidentally transpose or flip. Before fitting a
model, it is useful to verify:

1. **Dimensions:** do they match your expected grid resolution?
2. **Coordinates:** are x and y numeric and monotonic?
3. **Orientation:** does increasing y go "up" in plots, and does the covariate
   appear where you expect it geographically?
4. **Time alignment:** do covariate time slices cover the time range of the tag
   observations, and are interval start times correct?

A quick practical habit is to plot a known feature (e.g., coastline-aligned
bathymetry gradients) and confirm it looks correct in both your original data
and the prepared `admove_cov` object.

# Saving prepared data

To ensure reproducibility, save the prepared object to disk with `saveRDS()`:

```{r, eval=FALSE}
saveRDS(cov2, file = "prepared_cov.rds")
```

# Summary

This vignette described the covariate data structures expected by *admove* and
demonstrated how to convert common input types into the standard `admove_cov`
representation used internally by the package.

In practice, most workflows start from either:

- a raster product (GeoTIFF / NetCDF read into **terra** or **raster**), or
- a 3D array already close to `[x, y, time]`.

The key requirements are a regular grid, consistent orientation, and (ideally)
coordinate and time metadata stored as dimnames so covariate fields can be
aligned with tagging observations. Once prepared, standard methods such as
`summary()` and `plot()` can be used for quick inspection (e.g., `summary(cov2)`
or `plot(cov_r)`).

# References
