---
title: "Simulating tagging data"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Simulating tagging data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: admove.bib
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width = 8,
                      fig.height = 7,
                      dev = "png",
                      dev.args = list(bg = "white"))

knitr::knit_hooks$set(
  plot = function(x, options) {
    paste0(
      "<div style='background:white; padding:0.5rem; display:inline-block;'>",
      knitr::hook_plot_html(x, options),
      "</div>"
    )
  }
)

oldpar <- par(no.readonly = TRUE)
par(bg = "white")
```

# Introduction

This vignette demonstrates how to simulate tagging data using the **admove**
package. It covers both simple simulations using default settings and more
customised workflows that include spatial grids, environmental covariates,
model parameters, and release strategies for both conventional (mark-recapture)
and archival (data-storage) tags.

The simulation tools in **admove** are designed to support method development,
simulation studies, and testing of movement models in ecological applications.

**Outline:**

1. Quick simulation
2. Create grid
3. Environmental data
4. Parameters
5. Conventional tags
6. Archival tags
7. Simulate from fit
8. Summary
9. References

The examples below require the **admove** package:

```{r}
library(admove)
```

For reproducibility, we set a random seed:

```{r}
set.seed(123)
```

# Quick simulation

The fastest way to simulate tagging data is using the `sim_data()` function:

```{r}
sim <- sim_data()
```

This generates a complete simulated dataset. The returned object (class
`admove_sim`) contains:

- `sim$grid`: spatial grid
- `sim$cov`: simulated covariate fields
- `sim$par_sim`: simulation parameters
- `sim$tags`: simulated tagging data
- `sim$dat`: processed data for model fitting
- `sim$conf`, `sim$par`, `sim$map`: default configuration, parameters, and mappings

The simulation can be visualised with:

```{r}
plot(sim)
```

While convenient, `sim_data()` relies on default assumptions. Most of these can
be adjusted via its arguments (`?sim_data`). For full flexibility, the individual
components can be constructed manually, as shown below.

# Create a grid

The first step in a custom simulation is defining a spatial grid. The grid
determines the spatial domain, coordinate reference system (CRS), and resolution,
and is used to simulate covariate fields.

Here, we construct a grid resembling the eastern Pacific Ocean (similar to the
`skjepo` example dataset):

```{r}
xlim <- c(-150, -70)
ylim <- c(-30, 30)

idx <- c(1:63,65:78,81:94,97:110,113:126,129:141,145:154,161:169,177:183)

grid <- create_grid(xrange = xlim,
                    yrange = ylim,
                    cellsize = 5,
                    select = idx,
                    crs = 4326)
```

The index vector `idx` specifies selected grid cells, allowing reproducible
selection without interactive input. In practice, you may instead:

- include all cells (`select = FALSE`),
- interactively select cells (`select = 1`), or
- interactively remove cells (`select = 2`).

Visualise the grid:

```{r}
plot(grid, plot_land = TRUE)
```

# Covariate fields

Environmental covariates can be simulated using `sim_cov()`. By default, this
uses a Matérn spatial covariance structure and an AR(1) process in time:

```{r}
cov <- sim_cov(grid, nt = 9, rho_t = 0.4)
```

Here, nine temporal layers are generated with moderate temporal autocorrelation.

```{r}
plot(cov, plot_land = TRUE)
```

We also define a temporal reference:

```{r}
cov <- add_tref(cov, list(origin = as.Date("2020-01-01"),
                          units = "quarter"))
```

# Parameters

Next, we define the movement parameters.

- `alpha`: habitat preference (taxis)
- `beta`: diffusion

Both must be 3D arrays with dimensions:

1. number of knots
2. number of covariates
3. number of time periods

A simple example with constant diffusion and time-invariant preference:

```{r}
par <- list(
  alpha = array(c(0, 20, 10), dim = c(3, 1, 1)),
  beta  = array(log(5), dim = c(1, 1, 1))
)
```

Knots for the habitat preference function can be defined from covariate
quantiles:

```{r}
knots_tax <- matrix(
  round(quantile(as.numeric(cov),
                 c(0.05, 0.5, 0.95),
                 na.rm = TRUE), 2),
  3, 1
)

knots_dif <- matrix(0, 1, 1)
```

# Conventional (mark-recapture) tags

We simulate release locations:

```{r}
release_events <- sim_release_events(grid = grid)
```

```{r, fig.alt = "Release events"}
plot(grid, auto_layout = FALSE, labels = FALSE,
     plot_bg = FALSE, plot_land = TRUE)
points(release_events[,1], release_events[,2],
       col = "darkorange", pch = 16, cex = 2)
```

Now simulate tagging data:

```{r}
ctags_simulation <- sim_tags("c",
                            grid, cov, par,
                            n_tags = 100,
                            release_events = release_events,
                            knots_tax = knots_tax,
                            knots_dif = knots_dif,
                            tref = list(origin = "2020-01-01",
                                        units = "month"))
```

```{r}
plot(ctags_simulation$tags,
     leg_pos = "bottomright",
     plot_land = TRUE)
```

# Archival (data-storage) tags

Archival tags are simulated analogously:

```{r}
dtags_simulation <- sim_tags("d",
                            grid, cov, par,
                            n_tags = 5,
                            release_events = release_events,
                            knots_tax = knots_tax,
                            knots_dif = knots_dif,
                            tref = list(origin = "2020-01-01",
                                        units = "month"))
```

```{r}
plot(dtags_simulation$tags,
     leg_pos = "bottomright",
     plot_land = TRUE)
```

# Simulate from a fitted model

New datasets can also be simulated from an existing fitted model, such as
`skjepo$fit`:

```{r}
new_data <- sim_data(skjepo$fit, simulate_cov = FALSE)
```

Setting `simulate_cov = FALSE` reuses the original covariate fields, while
`TRUE` generates new ones.

```{r}
plot(new_data)
```

By default, release locations and times are random. For more control, users can
combine this approach with custom grids, covariates, and release strategies.

# Summary

This vignette demonstrated how to simulate tagging data in **admove**, ranging
from quick default simulations to fully customised workflows.

We showed how to:

- define spatial grids and domains,
- simulate environmental covariates with temporal structure,
- specify biologically meaningful movement parameters,
- generate both conventional and archival tagging data, and
- simulate new datasets from fitted models.

These tools allow users to construct realistic and flexible simulation setups
that reflect their study system. Carefully designed simulations are essential
for evaluating model performance, testing assumptions, and developing robust
movement models.

In practice, particular attention should be paid to consistency between spatial
and temporal scales, covariate structure, and parameter choices, as these
strongly influence the resulting movement dynamics and inference.


```{r, echo=FALSE}
par(oldpar)
```


# References
