---
title: "Spatial grids for admove"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Spatial grids for admove}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: admove.bib
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

This vignette provides more information about spatial grids in **admove**. In
particular, it demonstrates how to create and modify spatial grids using the
`create_grid()` function.

Although the Kalman filter (KF) engine does not necessarily require a grid
AUTHOR_PAPER, it is good practice to define a spatial grid based on the
spatial extent of the tagging data. The grid is required by the Continuous time
Markov-chain (CTMC) engine and is used by both engines for the spatial
prediction of taxis, advection, diffusion and movement rates and for
visualisation. The `create_grid()` function offers flexible functionality to
define custom grids and to manually include or exclude specific grid cells as
needed.

```{r}
library(admove)
library(sf)
```

The **sf** package is not a strict requirement of **admove**, but it provides a
lot of really useful functionality when working with spatial information, such
as spatial projections in **admove**.


# The most simple grid

The most simple default grid can be created by using `create_grid()` without arguments:

```{r}
grid <- create_grid()
```

This results in a unit grid with 100 cells:

```{r}
summary(grid)
```

which we can easily visualise with:

```{r, fig.alt = "The most simple spatial grid"}
plot(grid)
```

It is easy to change the cell size and make a coarser grid:

```{r, fig.alt = "Simple grid with coarse cellsize"}
grid <- create_grid(grid, cellsize = 0.2, plot = TRUE)
```

In a more realistic scenario, you might want to specify a grid in a specific
region given some x and y ranges. These could be longitude / latitude ranges or
correspond to any other units of a spatial projection.

```{r, fig.alt = "Using x and y ranges"}
grid <- create_grid(xrange = c(-150, -70),
                    yrange = c(-30, 35),
                    cellsize = 10,
                    plot = TRUE)
```

Note, that it is required to add spatial reference information (`sref`) to
define the coordinate reference system (CRS) and units. Only then land masses
can be plotted and the values of the arguments `xrange` and `yrange` are
defined. One way to add the CRS is as an argument to the `create_grid()`
function directly:

```{r, fig.alt = "Defining crs"}
grid <- create_grid(xrange = c(-15, 15),
                    yrange = c(30, 75),
                    cellsize = 5,
                    plot = TRUE,
                    crs = 4326,
                    plot_land = TRUE)
```

Here, we added the numeric code for the WGS 84 CRS in longitude/latitude and
consequently the landmasses are shown in the plot. Another way to add the
spatial reference information is after creating the grid:

```{r, fig.alt = "Using add_sref"}
## Land masses not plotted if spatial reference not defined
grid <- create_grid(xrange = c(-15, 15),
                    yrange = c(30, 75),
                    cellsize = 5,
                    plot = TRUE,
                    plot_land = TRUE)

## Add the spatial reference information
grid <- add_sref(grid, list(crs = 4326))

## Now the landmasses can be plotted
plot(grid, plot_land = TRUE)
```

When working with spatial projected grids, it might be nice to change units. For
example, the Albers projection

```{r}
albers_crs <- sf::st_crs(3035)
```

is in m by default, which can lead to large values:

```{r, fig.alt = "Albers leading to large values"}
grid <- create_grid(xrange = c(2000000, 5000000),
                    yrange = c(1500000, 5000000),
                    cellsize = 500000,
                    crs = albers_crs,
                    plot = TRUE,
                    plot_land = TRUE)
```


In this case, it might be nicer to work in km:

```{r, fig.alt = "Albers in km"}
grid <- create_grid(xrange = c(2000000, 5000000),
                    yrange = c(1500000, 5000000),
                    cellsize = 500000,
                    crs = albers_crs,
                    units = "km",
                    plot = TRUE,
                    plot_land = TRUE)
```

Or using an even smaller, custom scaling than m to km:

```{r, fig.alt = "Albers with custom units"}
grid <- create_grid(xrange = c(2000000, 5000000),
                    yrange = c(1500000, 5000000),
                    cellsize = 500000,
                    crs = albers_crs,
                    crs_scale = 1e-6,
                    plot = TRUE,
                    plot_land = TRUE)
```

# Selecting grid cells

Another useful functionality of `create_grid()` is that it allows to manually
select all cells to be included in the grid by setting `select = 1`. Any
non-selected cell will be `NA` and interpreted as a mask/boundary by **admove**,
i.e. excluded from the generator matrix.

```{r, eval=FALSE, echo=TRUE, fig.alt = "Selecting grid cells"}
grid <- create_grid(xrange = c(-150, -70),
                    yrange = c(-30, 35),
                    cellsize = 10,
                    select = 1,
                    crs = 4326,
                    plot_land = TRUE)
```

Specifying the CRS and setting `plot_land = TRUE` can help if the
selection is based on land/water boundaries.

When working in large spatial domains or many hundreds of fine grid cells it
might seem tedious to select all cells. Therefore, it is also possible to do the
opposite and de-select individual cells with `select = 2`.

```{r, eval=FALSE, echo=TRUE, fig.alt = "De-seelecting grid cells"}
grid <- create_grid(xrange = c(-150, -70),
                    yrange = c(-30, 35),
                    cellsize = 10,
                    select = 2,
                    crs = 4326,
                    plot_land = TRUE)
```

Unfortunately, this interactive functionality cannot be well demonstrated in a
vignette. Moreover, this manual selection might be a poor option in terms of
reproducibility or automatic scripts. Therefore, the argument `select` also
accepts a vector with the grid cell indices that should be selected, as for
example here to select cells in the Eastern Pacific Ocean:

```{r, fig.alt = "Selecting grid cells with indices"}
grid <- create_grid(xrange = c(-150, -70),
                    yrange = c(-30, 35),
                    cellsize = 10,
                    select = c(1:15,17:23,25:31,33:37,41:44),
                    crs = 4326,
                    plot = TRUE,
                    plot_land = TRUE)
```

Similarly, cells can be de-selected if negative indices are used:

```{r, fig.alt = "De-selecting grid cells with negative indices"}
grid <- create_grid(xrange = c(-150, -70),
                    yrange = c(-30, 35),
                    cellsize = 10,
                    select = -c(16,24,32,38:40,45:48),
                    crs = 4326,
                    plot = TRUE,
                    plot_land = TRUE)
```



# A spatial grid based on tags

Another convenient way to define a spatial grid with `create_grid()` is the feed
it other **admove** objects, such as tagging data or covariate fields. Let's
assume you have smoe mark-recapture tagging data available, for demonstration we
use the simulated example mark-recapture tags for skipjack tuna in the Eastern
Pacific Ocean included in **admove**:

```{r}
head(skjepo$ctags)
```

With one line, these tags can be converted into the expected format by
**admove**, check out `vignette("preparing-tags")` for more details on preparing
tagging data:

```{r}
ctags <- prep_tags(x = skjepo$ctags,
                   tag_type = "c",
                   names = c(t0 = "date_time",
                             x0 = "rel_lon",
                             y0 = "rel_lat",
                             t1 = "date_caught",
                             x1 = "recap_lon",
                             y1 = "recap_lat"),
                   date_origin = "1899-12-30",
                   sref = list(crs = 4326))

head(ctags)
```

Once they are processed, we can just define a default spatial grid based on them:

```{r, fig.alt = "Spatial grid based on tags"}
grid <- create_grid(ctags,
                    plot = TRUE,
                    plot_land = TRUE,
                    auto_layout = FALSE)

## plot tags on top of the spatial grid
plot(ctags, add = TRUE, auto_layout = FALSE)
```

The dimensions of the grid are defined based on the extreme positions in the
tagging data. If the cell size is not specified, it is defined to yield 100 grid
cells in total.


# A spatial grid based on covariate fields

Similarly, a spatial grid can be defined based on available **admove** covariate
fields. Let's use the covariate fields of the simulated example data for
Montagus harrier:

```{r, fig.alt = "Example covariate field for Montagus harrier"}
plot(montagus_harrier$cov[,,1,drop=FALSE])
```

As with the tagging data, we can just pass the covariate fields to
`create_grid()`:

```{r, fig.alt = "Spatial grid based on covariate fields"}
grid <- create_grid(montagus_harrier$cov,
                    plot = TRUE,
                    plot_land = TRUE,
                    auto_layout = FALSE)
```

If not specified, the cell size of the covariate field is taken over for
covariate fields. But, it would be easy enough to create a coarser cell size:

```{r, fig.alt = "Spatial grid based on covariate fields with coarser grid cells"}
grid <- create_grid(montagus_harrier$cov,
                    cellsize = 1,
                    plot = TRUE,
                    plot_land = TRUE,
                    auto_layout = FALSE)
```

Similarly, to tags and covariate fields, `create_grid()` accepts any other
`admove_*` classes, for example lists with simulated data (`admove_sim`) as
created by `sim_data()` or **admove**'s data list (`admove_data`) as created by
`setup_data()`.


# A spatial grid based on a spatial object

Another useful feature about `create_grid()` is that it accepts spatial objects
from other packages, such as spatial objects (of class `sf` or `sfc`) from R
package **sf** [@pebesma_simple_2018,@pebesma_spatial_2023] or `RasterLayer`
from the R package **raster** [@hijmans_raster_2025a]. For example we can use
part of the world map contained in **admove** to create a grid for Lake Eerie in
North America:

```{r, fig.alt = "Create spatial sf object"}
## get land from admove's internal data
land <- admove:::.get_land()

## cut to Northeast Atlantic
land_nea <- sf::st_crop(land, sf::st_bbox(
  c(xmin = -45, xmax = 35,
    ymin = 30,  ymax = 80),
  crs = sf::st_crs(4326)
))

## dissolve to a single geometry
land_nea <- sf::st_union(sf::st_geometry(land_nea))

## create proper sf object
land_nea <- sf::st_sf(geometry = land_nea, crs = 4326)

## plot the map
plot(sf::st_geometry(land_nea))
```

By default, this selects all grid cells on land:


```{r, fig.alt = "Creating grid based on sf object"}
grid <- create_grid(land_nea)
plot(grid, plot_land = TRUE)
```

By setting the argument `select = -1` we can turn the default selection around
and select every grid cells that is not on land:

```{r, fig.alt = "Creating grid based on sf object (inverse selection)"}
grid <- create_grid(land_nea, select = -1)
plot(grid, plot_land = TRUE)
```

If we are interested with a either starting selection and then continue choosing
cells by selecting or de-selecting, we can set `select 1` or  `select = 2`
with the default or inverse selection of cells inside polygons, respectively:

```{r, eval = FALSE, fig.alt = "Creating grid based on sf object (plus manual selection)"}
grid <- create_grid(land_nea, select = 1)
plot(grid, plot_land = TRUE)
```


# Summary

# References
