---
title: "Fine-scale movement modeling based on tagging data"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Fine-scale movement modeling based on tagging data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: admove.bib
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

This tutorial illustrates the application of the *admove* package to model
fine-scale movement based on tagging data.

**Outline:**

1. Prepare the data
2. Fit the model
3. Results
4. Advanced settings
5. Summary
6. References


The package is loaded into the R environment with:

```{r, eval=TRUE, echo=TRUE}
library(admove)
```

This introductory tutorial uses a simulated data set included in the package:
"skjepo". It is already available after loading the package and corresponds to a
list that includes the main simulation object (of class `admove_sim`), which
contains all required simulated data and can be used right away to fit the
model, as well as five individual data sets: `grid`, `cov`, `ctags`, `dtags`,
and a fitted object `fit`.

```{r, eval=TRUE, echo=TRUE}
str(skjepo, 1)
```

These data sets illustrate the typical structure of raw input data and
demonstrate how the functions in the *admove* package can be used to convert
them into the format required by the model as demonstrated in the next section.

# Prepare the data

The `skjepo` data list contains two main tagging data sets: one with information
on releases and recoveries of mark-recapture tags, and another with track data
from data-storage tags. These data sets can be prepared for analysis using the
built-in function `prep_tags()`, allowing users to specify which columns contain
the release and recapture times and locations, and automatically convert date
fields to the required format.

A key aspect of the preparation of tagging data and the use of this
function is the space and time reference information. The data set with the
mark-recapture tags contains information about the date and position when
individuals were released and recaptured:

```{r}
head(skjepo$ctags)
```

The time information might be a date of a numeric value corresponding to a date
or just relative time passed since an origin. Similarly, position information
might correspond to any coordinate reference system (CRS). This can either be
specified when using the `prep_tags()` function or afterwards. We show both ways
below.

Besides the data set and specifying the tag type ("c" for mark-re*c*apture, "d"
for *d*ata-storage, or "s" for mark-re*s*ight tags), a key argument for the
function is a named vector that links the required time ("t"), and positions
("x" and "y") with the column headers in the data set. Mark-recapture tags might
be available in two data formats: in the long format, similar to the
data-storage and mark-resight tags, with one time and two position columns, or
in the wide format with two time columns (containing the release ("t0") and
recapture time ("t1")) and four position columns ("x0" and "x1", and "y0" and
"y1"). For full details on how to use a function, consult its documentation. See
the "preparing-tags" vignette for more information on different tag formats and
how to convert tags between them.

```{r, eval=TRUE, echo=TRUE, fig.alt = "Mark-recapture and data-logging tags"}
## Prepare mark-recapture
ctags <- prep_tags(skjepo$ctags,
                   tag_type = "c",
                   names = c(t0 = "date_time", t1 = "date_caught",
                             x0 = "rel_lon", x1 = "recap_lon",
                             y0 = "rel_lat", y1 = "recap_lat"),
                   date_origin = "1899-12-30",
                   sref = list(crs = 4326),
                   tref = list(origin = as.POSIXct("2020-01-01", "UTC"),
                               units = "month"))

## Load dtags
dtags <- prep_tags(skjepo$dtags,
                   tag_type = "d",
                   names = c(t = "time", x = "mptlon", y = "mptlat"),
                   date_origin = "1899-12-30")

## Adjust the space reference information
dtags <- add_sref(dtags, list(crs = 4326))

## Adjust the time reference information
dtags <- add_tref(dtags, list(origin = as.POSIXct("2020-01-01", "UTC"),
                              units = "month"), shift_origin = TRUE)
```

As a next step, it is recommendable to check the space and time reference
information. All data components and objects in *admove* that contain
information about space and/or time include a `sref` (space reference
information) and a `tref` (time reference information) attribute. These can be
accessed with the `sref()` and `tref()` fuctions or investigated when printing
the summary of the any *admove* object:

```{r, eval=TRUE, echo=TRUE}
## Summarise the tagging information
summary(c(ctags, dtags))
```

The output provides some information about the tags and informs us that the WGS
84 [EPSG:4326] CRS was specified for the tags. The time reference information is
specified, here with weekly time steps and the date origin corresponding to
January 19th, 2020. This is the default behaviour of the `prep_tags()` function,
which does not infer space reference information by default but defines the
origin based on the first tag release time and the time step based on the
average life time of the tags.

Usually, it makes sense to specify a coordinate reference system and revise the
time reference information. This ensures that the space and time units and
scales are aligned between all *admove* components (e.g. covariate fields,
grids). We can specify time and space reference information and add them to the
tags with the following lines:

```{r, eval=TRUE, echo=TRUE}
## Define the target time reference information
tref_target <- create_tref(origin = as.POSIXct("2020-01-01", "UTC"),
                           units = "month")

## Define the target space reference information using Azimuthal Equidistant
## centered at (−110, 0)
sref_target <- create_sref(crs = "+proj=aeqd +lat_0=0 +lon_0=-110 +datum=WGS84 +units=m +no_defs",
                             units = "km")

## Adjust the reference info for the tags
ctags <- add_sref(ctags, sref_target, transform_crs = TRUE)
ctags <- add_tref(ctags, tref_target, shift_origin = TRUE)

dtags <- add_sref(dtags, sref_target, transform_crs = TRUE)
dtags <- add_tref(dtags, tref_target, shift_origin = TRUE)
```

Now, the tags have the required space and time reference information:

```{r, eval=TRUE, echo=TRUE}
summary(c(ctags, dtags))
```

Although the Kalman filter (KF) engine does not necessarily require a grid
AUTHOR_PAPER, it is good practice to define a spatial grid based on the
spatial extent of the tagging data. This grid is required by the Continuous
time Markov-chain (CTMC) engine and is used by both engines for the spatial
prediction of taxis, advection, diffusion and movement rates and for
visualisation. The `create_grid` function offers flexible functionality to
define custom grids and to manually include or exclude specific grid cells as
needed. See more information in the `vignette("spatial-grids")`.

Since manual selection of grid cells is not feasible in this automated vignette,
we use the grid provided in the skjepo data set, but modify it to create a
coarser resolution of 1000x1000km cells.

```{r, eval=TRUE, echo=TRUE, fig.alt = "Spatial grid"}
grid <- create_grid(skjepo$grid, cellsize = 750)
plot(grid, plot_land = TRUE)
```

Another important component of *admove* are covariate fields that taxis,
advection, and diffusion through habitat preference functions. The `prep_cov`
function formats this data into the structure required by the model. The
covariate fields can come from multiple sources and correspond to a range of
different formats, the `vignette("preparing-covariates")` gives more information
on the required format and how to convert data between different formats.

```{r, eval=TRUE, echo=TRUE, fig.alt = "First four covariate fields"}
## covariates
cov <- prep_cov(skjepo$cov)
plot_cov(cov[,,1:4], plot_land = TRUE)
```

It is not suprising if the covariate fields have different time reference
information. Here, for example, the covariate fields have the same date origin
as specified above (January 1st, 2020), but the time steps correspond to
quarters:

```{r, eval=TRUE, echo=TRUE, fig.alt = "First four covariate fields"}
tref(cov)
```

This is not a problem and the next functions would inform you about this
mismatch and provide tips on how to resolve them. Here we use one of the
potential ways by specifying the desired time reference information and setting
the argument `shift_tref = TRUE` in the next fuction that combines all data into
a single object.

```{r, eval=TRUE, echo=TRUE}
## combine and check data
dat <- setup_data(grid = grid,
                  cov = cov,
                  tags = c(ctags, dtags),
                  tref = tref_target,
                  shift_tref = TRUE)

```

The `default_conf` function generates a list of default configuration settings,
including flags that control which model functionalities are activated. It is
good practice to review this configuration list to ensure that the default
settings align with the goals and structure of your analysis.

```{r, eval=TRUE, echo=TRUE}
## Default configurations
conf <- default_conf(dat)
```

Similarly, the `default_par` function generates a list of default model parameters
and their initial values. While users can adjust the initial values as needed,
caution is required when modifying parameter dimensions. These dimensions must
be consistent with the input data (e.g., the number of environmental fields) and
the configuration settings (e.g., whether passive advection, taxis, or both are
used).

```{r, eval=TRUE, echo=TRUE}
## Default parameters
par <- default_par(dat, conf)
```
kappa is a scaling parameter that is not estimated but scales the spatial
gradient of the habitat preference function to a taxis rate. It carries the
dimensions L$^2$ L$^{-1}$ and might differ from 1 in the simulation study. To
get comparable parameter values between the simulation and estimation, it is
important to fix it to the value of the simulation in the estimation:

```{r, eval=TRUE, echo=TRUE}
par$logKappa <- skjepo$sim$par_sim$logKappa
```


With all necessary input data prepared, the model is now ready to be fitted
using *admove*.


# Fit the model

The model is fitted using the `admove` function. Depending on the model’s
complexity, the number of tags and the estimation engine chosen (KF or CTMC),
this step may take up to several minutes.

```{r, eval=TRUE, echo=TRUE}
## Fitting movement model
fit <- admove(dat, conf, par,
              verbose = TRUE)
```

The total run time for model construction, fitting, uncertainty estimation and
prediction was here (in minutes):

```{r, eval=TRUE, echo=TRUE}
sum(fit$times)
```

In addition to the input data lists, the returned object also includes the `obj`
(from RTMB) and `opt` (from the minimiser). Both contain the estimated parameter
values and other details relevant to the model fitting.

```{r, eval=TRUE, echo=TRUE}
fit$opt$par
```

# Results
The summary function can be used to get an overview of the model fit and
estimated parameters:

```{r, eval=TRUE, echo=TRUE}
summary(fit)
```

*admove* provides several functions for visualising model results. All plotting
functions start with `plot_.`, e.g. `plot_taxisis()` to plot the estimated taxis
as arrows on a map. To get a quick overview over the main results, we can use:

```{r, eval=TRUE, echo=TRUE, fig.alt = "Plot of results"}
plot(fit)
```

In this example, since the data are simulated and the true parameters are known,
we can use the `plot_compare` function to visualise both the estimated and
true habitat preferences and movement patterns.

```{r, eval=TRUE, echo=TRUE, fig.alt = "Simulated data and model predictions"}
plot_compare(sim = skjepo$sim,
             fit = fit,
             plot_land = TRUE)
```


# Advanced settings
There are several advanced settings in *admove* that will be covered in future
vignettes. However, this vignette highlights three important ones: (i) grouping
tags into release events to speed up model fitting, (ii) using CTMC engine as an
alternative to the KF engine, and (iii) map parameters to fix or exclude
them during estimation.

It is common for multiple tags to be released at the same location and time, in
particular for mark-recapture tags. Grouping these tags into release events can
significantly speed up the movement modeling, as the analysis is performed per
event rather than per individual tag. The potential loss in accuracy is likely
minimal and can be evaluated through sensitivity analyses, while the grouped
approach is recommended for the baseline scenario due to its efficiency. The
`use_release_event` function enables this grouping by assigning tags to release
events based on a specified spatial grid and time vector.

```{r}
tags <- use_release_events(c(ctags, dtags),
                           grid = create_grid(xrange = dat$xrange,
                                              yrange = dat$yrange,
                                              cellsize = 100),
                           time_cont = seq(dat$trange[1],
                                           dat$trange[2],
                                           1/4))
```

The release time and position of the mark-recapture tags has now been
overwritten to the values of the release events and the individual tags have
been combined to the individual release events:

```{r}
ctags <- get_ctags(tags)

str(split(ctags, ctags$id),1)
```

While the function could also identify release events for data-storage and
mark-resight tags using the argument `tag_types = c("d","s","c")`, this might
lead to unintended consequences when using the KF engine which includes an
update step.

By default, *admove* uses the Kalman filter approach as described in
AUTHOR_PAPER. However, it also supports an alternative method based on
the matrix exponential. You can easily switch between the two by setting the
`engine` argument in the configuration list. To use the matrix exponential
approach, simply set `engine = 2`.

```{r, eval=FALSE, echo=TRUE}
conf <- default_conf(dat)
conf$engine <- 2
```

Lastly, it may be useful to map parameters, for example, to estimate a single
value for both directions of passive advection. A list of default map
settings can be generated using the `default_map` function. The resulting
map can then be passed to `admove` to control which parameters are estimated,
fixed, or shared.

```{r, eval=FALSE, echo=TRUE}
map <- default_map(dat, conf, par)
```

## Summary
This vignette introduced the main features and workflow of the *admove* package for
estimating animal movement and habitat preferences from tagging data. Using a
simulated data set (`skjepo`), we demonstrated how to prepare the required inputs,
configure and fit the model, and visualise the results.

We covered the preparation of mark-recapture and data-storage tags
(`prep_tags()`), covariate fields (`prep_cov()`), and spatial grids
(`create_grid()`). These were then combined using `setup_data()` into a format
suitable for model fitting. Configuration and parameter initialisation were
handled through `default_conf()` and `default_par()`, with model fitting
performed using `admove()`. The `plot_fit()` and `plot_compare()` functions
allowed us to visualise results and compare estimated versus true movement and
habitat preference patterns.

Several advanced options were introduced, including grouping tags into release
events (`get_release_event()`), switching between the two currently implemented
engines for inferences: KF and CTMC engines (`engine` argument), and map
parameters (`default_map()`) to simplify or constrain estimation.

This basic example provides a foundation for applying *admove* to real tagging
data. More details about the methodology are described in AUTHOR_PAPER
and in the other vignettes (get an overview with `browseVignettes("admove")`).
Further details about functions and their arguments can be found in the help
files of the functions (`help()` or `?`, where the dots refer to any
function of the package).


# References
